Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-13 manuscript/manuscriptRW.Rmd
substitude rates per levles with ingrowth and added a new paragraph in research recomendations

Most of the models overestimate ingrowth rates, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth rates
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

@@

The three models with the largest overestimation of the proportion of
ingrowth at the 7 cm threshold featured the highest mortality between
the 7 and 10 cm thresholds (Figure \@ref(fig:mortality13)A). The other
models that overestimated ingrowth had mortalities lower than the
expected self-thinning ratio with the exception of LandClim (Figure
\@ref(fig:mortality13)B). It is noteworthy that most of the models that
featured a low ratio of ingrowth between 7 and 10 cm (i.e., ForCEEPS,
ForClim 1, ForClim 11 and iLand) had only a small overestimation of
ingrowth at the 7 cm threshold (Figure \@ref(fig:mortality11)). aDGVM2
was the only model that underestimated ingrowth at the 7 cm threshold
while featuring a low mortality between the two thresholds (Figure
\@ref(fig:mortality13)C), and only a few models were close to the
correct ingrowth rate at 7 cm while also being close to the Reineke line
(SIBYLA, Landis-II; cf. Figure \@ref(fig:mortality13)D; and 4C as well
as LandClim).

@@

Lastly, some models that had a low number of individuals at 7 cm
threshold also had a smaller decrease of ingrowth between 7 and 10 cm
(ratio 7-10 cm) (Figure S5). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S5). It is noteworthy that there is a relationship
between the ingrowth rate at 7 cm and the mortality rate (Table S4):
most of the models showed a positive effect (higher ingrowth at 7 cm is
coupled to an increase of the mortality between 7 and 10 cm). While some
models showed a negative effect, this linear trend was not pronounced
and not always significant regarding the slope.

@@

## Most of the models overestimated tree ingrowth rates

Most of the models overestimated tree ingrowth rates. This has
potentially far-reaching implications, e.g., regarding biomass (and thus
carbon) turnover, with a potential overestimation of the capacity for
forest carbon sequestration [@pugh2019; @pan2011]. Yet, we focused on
one specific stage, i.e., ingrowth into the 7 and 10 cm diameter
classes. Trees at this size contribute little to carbon sequestration,
and if the excess ingrowth rate at this stage is compensated -- in the
models -- soon thereafter by higher mortality, overall simulated
ingrowth may still yield trustworthy results at the level of stand
structure and related forest attributes such as basal area or biomass.

@@

Increasing complexity in establishment modules has been motivated by
better process understanding and enhancing of model accuracy
[@bugmann2022; @bugmann2019]. However, more complex models do not
necessarily lead to better projections and rather to a reduced
transparency and predictive power [@franklin2020]. The results from this
study showed that increasing complexity in the establishment modules did
not represent an improvement in the projections of ingrowth levels,
species composition or mortality at early tree stages. Thus, the
question what level of detail one has to use to model establishment
processes remains open [@konig_tree_2022; @bugmann2022].

@@

Competition for light as a strong filter for tree establishment has been
widely documented [@messier1999; @collet2006; @berdanier2016], but
models did not reproduce this expectation. However, it is difficult to
measure light availability at large spatial and temporal scales. We used
total stand basal area as a proxy for light availability [cf.
@schmid2021]. We found pronounced differences in the stand basal area
ranges simulated by the models, but also between the models and the
observations. This made not possible to evaluate the establishment for
the extremes of the stand density ranges in some models. For example,
ingrowth rates at low stand densities are relevant to assess how well
forests are recovering e.g., after gap creation due to disturbance
[@seidl2022; @grubb1977]. At the level of the simulated one-hectare
samples, average basal area is typically not very low as long as no
larger disturbances occur, which was explicitly excluded in our
simulation protocol. Towards the other end of the spectrum, i.e., with
increasing basal area, it would be reasonable to expect that with
increasing stand basal area, the ingrowth of the different species would
become sparser and drop out entirely under low-light conditions
[@klopcic2012; @zell2019]. However, few models showed this trend, thus
indicating that the relation between ingrowth and light availability is
not yet captured correctly in most models.

There were pronounced variations how the main tree species were
represented along the environmental gradients, in particular the
dominant species Abies alba, Fagus sylvatica and Picea abies. The
ingrowth rates were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
[@klopcic2015]. Most models captured this low abundance, which is also
found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

@@

The EuFoRIa data as used in this study are unique and well suited to
better understand tree establishment. However, three aspects of these
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a equilibrium state, but we cannot assess
how close the forests included in the EuFoRIa dataset were to such a
state. The data were collected in forest reserves where no management
has taken place for long periods of time. This makes our assumption of
an equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth rates were higher than the
observed ones. Second, the data were collected for ingrowth above a 7 cm
threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand some aspects of tree establishment, which in its entirety
often comprises a rather long period since seed production [@price2001].
However, many environmental constraints are acting on young trees
[@käber2021] that we were unable to assess. Yet, ingrowth data with
lower thresholds are simply not available in a harmonized manner across
large environmental gradients. Third, the empirical data were collected
from rather small plots, while we sampled simulated ingrowth rates from
1 ha areas, which may lead to an incorrect representation of space. Even
though this strategy is not ideal, it represents a common challenge when
harmonizing diverse data sources originating from varied sampling
strategies [@portier2022]. It would have been extremely challenging for
such a variety of models to follow a protocol where the spatial sampling
size was different at each of the 200 sites and it would have introduce
additional uncertainty into the results. We can not forget here that the
observed data is just a snapshot of an extremely complex and stochastic
reality, which might be the explanation of e.g. lack of clear patterns
when confronted with gradients of climater water balance and degree
days, while the models outputs are the realizations of the definitions
of the regeneration niche.

@@

@@

Models of forest dynamics are important tools in science and decision
support, and how tree establishment is defined has strong implications
for simulated forest properties. The 15 models and variants used here
are facing similar challenges when representing tree establishment. They
generally overestimate tree establishment rates, and the simulated
establishment niche is not always captured accurately as a function of
the biotic (light) and abiotic (temperature and moisture) factors.

>>>>

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

@@

The three models with the largest overestimation of the proportion of
ingrowth at the 7 cm threshold featured the highest mortality between
the 7 and 10 cm thresholds (Figure \@ref(fig:mortality13)A). The other
models that overestimated ingrowth had mortalities lower than the
expected self-thinning ratio with the exception of LandClim (Figure
\@ref(fig:mortality13)B). It is noteworthy that most of the models that
featured a low ratio of ingrowth between 7 and 10 cm (i.e., ForCEEPS,
ForClim 1, ForClim 11 and iLand) had only a small overestimation of
ingrowth at the 7 cm threshold (Figure \@ref(fig:mortality11)). aDGVM2
was the only model that underestimated ingrowth at the 7 cm threshold
while featuring a low mortality between the two thresholds (Figure
\@ref(fig:mortality13)C), and only a few models were close to the
correct ingrowth levels at 7 cm while also being close to the Reineke
line (SIBYLA, Landis-II; cf. Figure \@ref(fig:mortality13)D; and 4C as
well as LandClim).

@@

Lastly, some models that had a low number of individuals at 7 cm
threshold also had a smaller decrease of ingrowth between 7 and 10 cm
(ratio 7-10 cm) (Figure S5). In these models, competition and
self-thinning are either not pronounced or must have occurred before the
species had reached 7 cm. However, this pattern was not consistent
across models (Figure S5). It is noteworthy that there is a relationship
between the ingrowth levels at 7 cm and the mortality rate (Table S4):
most of the models showed a positive effect (higher ingrowth at 7 cm is
coupled to an increase of the mortality between 7 and 10 cm). While some
models showed a negative effect, this linear trend was not pronounced
and not always significant regarding the slope.

@@

## Most of the models overestimated tree ingrowth levels

Most of the models overestimated tree ingrowth levels. This has
potentially far-reaching implications, e.g., regarding biomass (and thus
carbon) turnover, with a potential overestimation of the capacity for
forest carbon sequestration [@pugh2019; @pan2011]. Yet, we focused on
one specific stage, i.e., ingrowth into the 7 and 10 cm diameter
classes. Trees at this size contribute little to carbon sequestration,
and if the excess ingrowth at this stage is compensated -- in the models
-- soon thereafter by higher mortality, overall simulated ingrowth may
still yield trustworthy results at the level of stand structure and
related forest attributes such as basal area or biomass.

@@

The results from this study showed that increasing complexity in the
establishment modules did not represent an improvement in the
projections of ingrowth levels, species composition or mortality at
early tree stages. Increasing complexity in establishment modules has
been motivated by better process understanding and enhancing of model
accuracy [@bugmann2022; @bugmann2019]. However, more complex models do
not necessarily lead to better projections and rather to a reduced
transparency and predictive power [@franklin2020]. Thus, the question
what level of detail one has to use to model establishment processes
remains open.[@konig_tree_2022; @bugmann2022].

@@

Competition for light as a strong filter for tree establishment has been
widely documented [@messier1999; @collet2006; @berdanier2016], but
models did not reproduce this expectation. However, it is difficult to
measure light availability at large spatial and temporal scales. We used
total stand basal area as a proxy for light availability [cf.
@schmid2021]. We found pronounced differences in the stand basal area
ranges simulated by the models, but also between the models and the
observations. This made not possible to evaluate the establishment for
the extremes of the stand density ranges in some models. For example,
ingrowth levels at low stand densities are relevant to assess how well
forests are recovering e.g., after gap creation due to disturbance
[@seidl2022; @grubb1977]. At the level of the simulated one-hectare
samples, average basal area is typically not very low as long as no
larger disturbances occur, which was explicitly excluded in our
simulation protocol. Towards the other end of the spectrum, i.e., with
increasing basal area, it would be reasonable to expect that with
increasing stand basal area, the ingrowth of the different species would
become sparser and drop out entirely under low-light conditions
[@klopcic2012; @zell2019]. However, few models showed this trend, thus
indicating that the relation between ingrowth and light availability is
not yet captured correctly in most models. On the other hand several of
the models with an increase of ingrowth with increasing basal area also
include the feedback between seed production of mature trees and
ingrowth. Thus it seems that the light competition does not sufficiently
compensate the increased seed availability with higher basal area.

There were pronounced variations how the main tree species were
represented along the environmental gradients, in particular the
dominant species Abies alba, Fagus sylvatica and Picea abies. The
ingrowth levels were sampled from simulations in the equilibrium, and in
this state it is expected that non-climax species such as *Pinus
sylvestris* or *Quercus* would be of minor importance, or absent
[@klopcic2015]. Most models captured this low abundance, which is also
found in the empirical data. Thus, although the broad patterns are
matched by many models, improvements in the quantification of the
regeneration niche of the species are needed, but this cannot be done in
the absence of robust datasets.

@@

The EuFoRIa data as used in this study are unique and well suited to
better understand tree establishment. However, three aspects of these
data may also represent considerable limitations. First, we made a
comparison of tree ingrowth in a equilibrium state, but we cannot assess
how close the forests included in the EuFoRIa dataset were to such a
state. The data were collected in forest reserves where no management
has taken place for long periods of time. This makes our assumption of
an equilibrium between forest properties and environmental drivers more
reasonable than it might appear at first sight. However, the equilibrium
assumption may explain why simulated ingrowth levels were higher than
the observed ones. Second, the data were collected for ingrowth above a
7 cm threshold, thus limiting the assessment of tree establishment to a
holistic viewpoint. This constituted a hard limit based on which we can
understand some aspects of tree establishment, which in its entirety
often comprises a rather long period since seed production [@price2001].
However, many environmental constraints are acting on young trees
[@käber2021] that we were unable to assess. Yet, ingrowth data with
lower thresholds are simply not available in a harmonized manner across
large environmental gradients. Third, the empirical data were collected
from rather small plots, while we sampled simulated ingrowth levels from
1 ha areas, which may lead to an incorrect representation of space. Even
though this strategy is not ideal, it represents a common challenge when
harmonizing diverse data sources originating from varied sampling
strategies [@portier2022]. It would have been extremely challenging for
such a variety of models to follow a protocol where the spatial sampling
size was different at each of the 200 sites and it would have introduce
additional uncertainty into the results. We can not forget here that the
observed data is just a snapshot of an extremely complex and stochastic
reality, which might be the explanation of e.g. lack of clear patterns
when confronted with gradients of climate water balance and degree days,
while the models outputs are the realizations of the definitions of the
regeneration niche.

@@

Exercises like the one presented here where the models are run in "blind
flight" without the possibility of being tuned towards capturing the
observed values should be repeated. Such a benchmarking exercises should
next focus in different aspects such as specific model traits and
ecological formulations to better understand the implications of the
model assumptions. Furthermore, the inclusion in this exercises with
different scales, type, populations structures, tree establishment
modules and complexity ensure a wider benefit to the whole modeling
community.

@@

Models of forest dynamics are important tools in science and decision
support, and how tree establishment is defined has strong implications
for simulated forest properties. The 15 models and variants used here
are facing similar challenges when representing tree establishment. They
generally overestimate tree establishment levels, and the simulated
establishment niche is not always captured accurately as a function of
the biotic (light) and abiotic (temperature and moisture) factors.
